Gravitational Wave Detection:\\ Event GW170814 and the performance of matched filtering and $\chi^2$-consistency testing in the presence of Gaussian noise, non-Gaussianly distributed noise, and transient glitches

### 2020-09-07 paper2.tex

\doublespacing
\begin{document}

\title{Gravitational Wave Detection:\\
Event GW170814 and the performance of matched filtering and $\chi^2$-consistency testing in the presence of Gaussian noise, non-Gaussianly distributed noise, and transient glitches}
\author{Vinh Nguyen}
\email[Mentored by Dr. Eric Myers – Department of Physics and Astronomy, State University of New York at New Paltz]{}
\affiliation{2020 Pioneer Academics Program \\ Klein Oak High School, Spring, TX 77379}
\date{July 29, 2020}
\begin{abstract}
The detection of gravitational wave signals of astrophysical origins is particularly challenging because the signals are always contaminated by noise. Two of most prominent and efficient detection techniques currently used in Virgo/LIGO search pipelines are matched filtering and $\chi^2$-consistency testing. This paper investigated the performance of these two detection techniques in event GW170814 as well as in the presence of both consistently distributed noise and transient glitches. It was found that matched filtering performs best in the presence of perfect or close-perfect Gaussian noise; it can also mistake glitches for real gravitational waves if certain types of glitches are present. Additionally, it was found that $\chi^2$-consistency testing has little effects in cases where matched filtering had previously been unsuccessful. Finally, it was shown that the use of $\chi^2$-consistency testing is crucial to detecting and identifying gravitational waves in the presence of \textit{Extremely Loud} and \textit{Blip} glitches.
\end{abstract}

\maketitle
\section{Introduction}
The existence of gravitational waves was first predicted by Albert Einstein in 1916 as a direct result of his Theory of General Relativity. Since then, gravitational wave science and detection have been a leading scientific field of interest for not just physicists, but also astronomers, geologists, material scientists, and engineers as well. With the first confirmed detection of gravitational waves from a binary black hole merger in 2015 (event GW150914), the prospects of gravitational wave science and its applications to other fields have only become brighter. As of July 2020, there exists four primary fully functioning gravitational wave detectors – the LIGO Hanford Detector in Washington, the LIGO Livingston Detector in Louisiana, the Virgo Observatory in Italy, and the KAGRA detector in Japan.

@@

This paper uses real Virgo/LIGO data queried from the Gravitational Wave Open Science Center (GWOSC). GWOSC is a collaborative effort between the LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration to provide the public with gravitational wave data from the two LIGO facilities (Hanford, WA and Livingston, LA) as well the Virgo Observatory in Cascina, Italy. The structure of this paper is as follows. \textit{Section \ref{background}} provides detailed background information necessary for the methods undertaken in this paper. Specifically, \textit{sub-section \ref{gw and ligo}} discusses gravitational waves, its detection and current detector technology while \textit{sub-section \ref{data anal}} addresses data analysis details such as  the Fourier transform, noise characteristics, matched filtering and $\chi^2$-consistency testing. \textit{Section \ref{method}} gives a detailed description of this paper's methodology and the steps undertaken. \textit{Section \ref{results}} provides the paper's preliminary results in addition to data analysis. \textit{Section \ref{conclusions}} presents the study's conclusions and proposals for possible future research directions. Finally, \textit{section \ref{summary}} gives a brief overall summary of this paper's content.

\section{Background} \label{background}
\subsection{Gravitational waves and the Laser Interferometer Gravitational-Wave Observatory (LIGO)} \label{gw and ligo}
In 1915, Albert Einstein first published his theory of general relativity, describing space and time as one four-dimensional concept called \textit{``space-time''} \cite{maggiore_2008}, In this theory, he describes the force of gravity not as an inherent property that a massive body possesses, but rather as a byproduct of curved space-time. To put this more simply, we can say that a body with mass warps space-time, thereby creating a gravitational field surrounding it, affecting other bodies \cite{carlip}. To understand this idea, we can make use of an analogy: suppose that the fabric of space-time is a rubber sheet, and that any object with mass on it warps the surrounding area (FIG. 1.), hence creating a gravitational field. It follows from Einstein's Theory of Relativity that rapid changes in the curvature of space-time would produce ripples, called \textit{gravitational waves} (abbreviated as GW), that propagate outward at the speed of light  \cite{JSTORLIGO}.

@@

For gravitational waves passing through the test masses, the fractional difference in arm length $\frac{\Delta L}{L}$ is the strain $h$ (a quantity that measure the `strength' of gravitational waves) of the waves. Mathematically, the strain $h(t) = \frac{\Delta L(t)}{L}$ can also be expressed as a linear combination of its cross and plus polarizations, as follows \cite{JSTORLIGO}:
\begin{equation}
h(t) = F_\times h_\times (t) + F_+ h_+ (t)
\end{equation}

@@

Since we are working with discrete-valued sequences, performing a DFT would give rise to what we call \textit{spectral leakages}. When the Fourier Transform is applied onto a finite, non-periodic data segment (i.e.\ Virgo/LIGO data), false frequency amplitudes might occur. The reason for this is that since the transformation is discrete in nature, frequencies that lies in between or across multiple frequency intervals can falsely create high amplitudes. Thus, to minimize this, we always apply a windowing function to data – tapering the data segment's ends, making it periodic – whenever the use of the DFT is required (note that windowing does not decrease spectral leakage, but rather re-distributes the leakage effect so that it causes the least harm). In this paper, I used a computer algorithm that implements the Discrete Fourier Transform called the Fast Fourier Transform (FFT) which is included within the Python PyCBC library \cite{numpy}.

@@

\subsubsection{Noise characteristics}
Being able to reach a strain sensitivity of $10^{-23}/\sqrt{\text{Hz}}$ \cite{sensitivity} means Virgo/LIGO's interferometers are extremely sensitive to any changes in the arm lengths – which can be caused by distorted space-time due to gravitational waves, or by vibrations that move the test masses. Thus, recorded signal $s(t)$ includes not only gravitational waves, but also unwanted vibrations from other sources as well \cite{ultimate}. Any unwanted noise, even on the smallest of scales, can still significantly interfere with data recorded by Virgo/LIGO's interferometers. Virgo/LIGO noise comes from many sources, but some examples include seismic noise due to motions of the Earth's surface, thermal noise at the molecular level affecting the test masses, and quantum shot noise at high frequencies caused by the quantum nature of light \cite{blair_howell_ju_zhao_2012}.

In order to extract buried gravitational wave signals from noise, we must first understand the statistical properties of the noise. Noise is said to be stationary if its statistical properties do not significantly change over time. Another important noise characteristic is its Gaussianity. Given \textit{`Gaussian'} noise with a time-series function $n(t)$, both the real and imaginary components of its Fourier Transform $\widetilde{n}(f)$ follow the normal distribution \cite{yamamoto}; that is, $\Re[\widetilde{n}(f)]$ and  $\Im{[\widetilde{n}(f)]}$ have a probability density function (PDF) of:
\begin{equation}
    p(x) = \frac{1}{\sigma\sqrt{2\pi}}\exp{\left(-\frac{1}{2}\frac{(x-\mu)^2}{\sigma^2}\right)}
\end{equation}
where $\sigma^2$ is the sample variance and the mean $\mu = 0$. By the Central Limit Theorem in statistics, a set of random noise sample naturally tends to the Gaussian distribution if sufficiently large data samples were taken \cite{jaranowski2007gravitationalwave}. Nonetheless, real Virgo/LIGO noise can still deviate from this theoretical normality. Aside from the Gaussian distribution, noise can also follow a variety of different probability density distributions such as the Poisson distribution or the Laplace distribution. Although it is not common in real-life Virgo/LIGO operations that these noise models are present, this paper still considers the case of Laplacian noise for theoretical analysis purposes. Furthermore, there also exists abrupt, transient noise sources of, sometimes, unknown origins called glitches \cite{ultimate}. Glitches can be grouped into a variety of classifications according Gravity Spy's \textit{`Field Guide'} \cite{gravityspy}. In this paper, we only primarily discuss five types of glitches: \textit{Extremely Loud}, \textit{Scattered Light}, \textit{Power Line}, \textit{Whistle}, and \textit{Blip}. \textit{Extremely Loud} is a broad classification for major detector disturbances such as photodiode saturation, causing loud and high energy transient noise bursts. \textit{Scattered Light} glitches are caused by malfunctions in the optical mechanisms of Virgo/LIGO; they look like upward humps and are particularly harmful in searching for binary inspirals \cite{gravityspy}. The third classification of glitches is called \textit{Power Line}; \textit{Power Line} glitches are caused by electrical equipments such as air compressors that use alternating current (AC). The fourth classification is called \textit{Whistle} glitches; they are usually reverse V-shaped and are caused by radio signals \cite{gravityspy}. Finally, \textit{Blip} glitches are very short duration ($\approx 40$ms) noise bursts at frequencies between 30 and 500 Hz. As of currently, there are no known causes and thus \textit{Blip} glitches are the most damaging to the detection of gravitational waves \cite{gravityspy}. It is worth pointing out that over short time intervals, though, Virgo/LIGO noise can still be reasonably approximated to be Gaussian and stationary \cite{collaboration2019open}.

\begin{widetext}
\begin{figure}[t]
\includegraphics[width = .9\textwidth]{gaussian -21 template 1.png}
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Gaussian noise scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 20.0$}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 gaussian zoomed.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Gaussian noise scaled down by a factor of $10^{-21}$. A strong dip toward unity is clearly seen at the expected time $t=35s$.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 gaussian.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Gaussian noise scaled down by a factor of $10^{-21}$, zoomed in to just the two-second interval between the time of expected signal.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{laplacian b=1 template 1.png}
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Laplacian noise with $b=1$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 12.6$.}

@@

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=1 zoomed.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=1$ scaled down by a factor of $10^{-21}$. A strong dip toward unity is seen at time $t = 35.0$s with climbing values just around it.}
\centering
\end{figure} 
\end{widetext}

@@

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=1.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=1$ scaled down by a factor of $10^{-21}$. The graph is zoomed in to show just the two-second time interview around the time of expected signal. A strong dip toward unity is seen at time $t = 35.0$s.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{laplacian b=2 template 1.png}
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Laplacian noise with $b=2$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 6.1$.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=2 zoomed.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=2$ scaled down by a factor of $10^{-21}$. A small dip toward unity is seen at time $t = 35.0$s.}
\centering
\end{figure} 
\end{widetext}
\subsubsection{Matched filtering}
The primary goal of Virgo/LIGO data analysis is to determine if there is gravitational wave signal buried in the noise. There are many ways that this can be accomplished; however, the most efficient and effective method is called \textit{matched filtering}, especially in cases where noise is both Gaussian and stationary \cite{helstrom_1975}. Matched filtering allows the recorded data stream to be searched for the presence of a known gravitational wave template with varying arrival time and phase offset \cite{surf}. Matched filtering requires three components: the recorded data stream $s(t)$, the \textit{power spectral density (PSD)} function $S_n(t)$ of the noise, and a known waveform template $h(t)$ of the gravitational wave. (Note that the PSD is essentially a measure of how power is distributed at different frequencies of the signal; detailed mathematical descriptions can be found in \cite{psd}.) Given these quantities, the complex-valued output of matched filtering at time $t=t_0$ can be computed as follows \cite{findchirp}:
\begin{equation}
    z(t_0) = 4\int_0^\infty\frac{\widetilde{s}(f)\widetilde{h}^*(f)}{S_n(f)}e^{2\pi\iu ft_0}df
\end{equation}
where $\widetilde{h}^*(f)$ denotes the complex conjugate of the Fourier Transform of the template strain $h(t)$. Notice that the existence of $S_n(f)$ in the denominator essentially down-weighs the frequencies where the noise is loud and does the opposite where the noise is weak \cite{ultimate}. Given the complex matched filter output $z(t)$, we can compute a test statistic value $\rho$ called the \textit{signal-to-noise ratio} as follows \cite{findchirp}: 
\begin{equation}
   \rho (t) = \frac{|z(t)|}{\sigma}
\end{equation}
where $\sigma$ is a normalization factor is computed from \cite{findchirp}
\begin{equation}
   \sigma^2 =  4\int_0^\infty\frac{|\widetilde{h}(f)|^2}{S_n(f)}df
\end{equation}

@@

In the presence of non-Gaussian and/or non-stationary noise and/or glitches, matched filtering does not perform as well as expected since the filter itself is derived from Gaussian and stationary noise. A detailed mathematical derivation of the matched filter from a stationary, Gaussian noise model can be found in \textit{Brown (2007)} \cite{brown2007searching}.

There are several ways of incorporating matched filtering into a computer program that analyzes LIGO's signal. In this paper, I only make use of the matched filtering algorithm implemented in the open-source Python software package PyCBC \cite{pycbc}. This algorithm is based on the FINDCHIRP algorithm presented above and in \cite{findchirp}.
\begin{widetext}
\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=2.png}
\caption{This figure shows the resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=2$ scaled down by a factor of $10^{-21}$. The graph is zoomed in to show just the two-second time interview around the time of expected signal. A relatively weak dip toward unity is seen at time $t = 35.0$s.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{laplacian b=3 template 1.png}
\caption{This figure shows the resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Laplacian noise with $b=3$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 21.7$s with SNR $\rho = 4.6$.}
\centering
\end{figure}
\end{widetext}

>>>>

\begin{document}

\title{Gravitational Wave Detection:\\
Event GW170814 and the performance of matched filtering and $\chi^2$-consistency testing in the presence of Gaussian noise, non-Gaussianly distributed noise, and transient glitches}
\author{Vinh Nguyen}
\email[Mentored by Dr. Eric Myers – Department of Physics and Astronomy, State University of New York at New Paltz]{}
\affiliation{Klein Oak High School, Spring, TX 77379}
\date{July 29, 2020}
\begin{abstract}
The detection of gravitational wave signals of astrophysical origin is particularly challenging because the signals are always contaminated by noise. Two of most prominent and efficient detection techniques currently used in Virgo/LIGO search pipelines are matched filtering and $\chi^2$-consistency testing. This paper investigated the performance of these two detection techniques in event GW170814 as well as in the presence of both consistently distributed noise and transient glitches. It was found that matched filtering performs best in the presence of perfect or close-perfect Gaussian noise; it can also mistake glitches for real gravitational waves if certain types of glitches are present. Additionally, it was found that $\chi^2$-consistency testing has little effects in cases where matched filtering had previously been unsuccessful. Finally, it was shown that the use of $\chi^2$-consistency testing is crucial to detecting and identifying gravitational waves in the presence of \textit{Extremely Loud} and \textit{Blip} glitches.
\end{abstract}

\maketitle
\section{Introduction}
The existence of gravitational waves was first predicted by Albert Einstein in 1916 as a direct result of his Theory of General Relativity. Since then, gravitational wave science and detection have been a leading scientific field of interest for not just physicists, but also astronomers, geologists, material scientists, and engineers. With the first confirmed detection of gravitational waves from a binary black hole merger in 2015 (event GW150914), the prospects of gravitational wave science and its applications to other fields have only become brighter. As of July 2020, there exist five primary fully functioning gravitational wave detectors – the LIGO Hanford Detector in Washington, the LIGO Livingston Detector in Louisiana, the Virgo Observatory in Italy, the GEO600 detector in Germany and the KAGRA detector in Japan.

@@

This paper uses real Virgo/LIGO data queried from the Gravitational Wave Open Science Center (GWOSC). GWOSC is a collaborative effort between the LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration to provide the public with gravitational wave data from the two LIGO facilities (Hanford, WA and Livingston, LA) as well the Virgo Observatory in Cascina, Italy. The structure of this paper is as follows. \textit{Section \ref{background}} provides detailed background information necessary for the methods undertaken in this paper. Specifically, \textit{sub-section \ref{gw and ligo}} discusses gravitational waves, their detection and current detector technology while \textit{sub-section \ref{data anal}} addresses data analysis details such as  the Fourier transform, noise characteristics, matched filtering and $\chi^2$-consistency testing. \textit{Section \ref{method}} gives a detailed description of this paper's methodology and the steps undertaken. \textit{Section \ref{results}} provides the paper's results in addition to data analysis. \textit{Section \ref{conclusions}} presents the study's conclusions and proposals for possible future research directions. Finally, \textit{section \ref{summary}} gives a brief overall summary of this paper's content. The source code in LaTeX for this paper can be found on the Github repository \href{https://github.com/VinhNguyen313/LIGO-and-Gravitational-Waves-Project}{\textbf{\textit{here}}}.

\section{Background} \label{background}
\subsection{Gravitational waves and the Laser Interferometer Gravitational-Wave Observatory (LIGO)} \label{gw and ligo}
\subsubsection{Gravitational waves and their detection}
In 1915, Albert Einstein first published his theory of general relativity, describing space and time as one four-dimensional concept called \textit{``space-time''} \cite{maggiore_2008}, In this theory, he describes the force of gravity not as an inherent property that a massive body possesses, but rather as a byproduct of curved space-time. To put this more simply, we can say that a body with mass warps space-time, thereby creating a gravitational field surrounding it, affecting other bodies \cite{carlip}. To understand this idea, we can make use of an analogy: suppose that the fabric of space-time is represented by a rubber sheet, and that any object with mass on it warps the surrounding area (FIG. 1), hence creating a gravitational field. It follows from Einstein's Theory of Relativity that rapid changes in the curvature of space-time (specifically through a changing quadrupole mass distribution) would produce ripples, called \textit{gravitational waves} (abbreviated as GW), that propagate outward at the speed of light  \cite{JSTORLIGO}. On a astronomical scale, sources of gravitational waves can be classified into three distinct classes \cite{blair_howell_ju_zhao_2012}:

\begin{enumerate}
\item Burst Sources

Examples of burst sources include the final coalescence of compact binary star systems, and asymmetric supernovae. The binary coalescence events can consist of binary neutron stars, binary black holes, or neutron star–black hole
binaries. Note that a burst signal consists of a very short single event; thereby it usually has a broad bandwidth, roughly determined by the
reciprocal of the event duration \cite{blair_howell_ju_zhao_2012}.
\item Narrow Band Sources

Examples of narrow band sources include the rotation of single non-axisymmetric stars such as pulsars, accreting neutron stars, and binary star systems far from coalescence. Gravitational waves from these sources are also generally weaker than those originating from burst sources \cite{blair_howell_ju_zhao_2012}.
\item Stochastic Background Sources

Gravitational waves from these sources are remnants from the early universe after the Big Bang. They can be produced from the combined effects of weak periodic sources throughout the Galaxy, from a large population of burst sources at very large distances, or from the other cosmological events mentioned above.
\end{enumerate}

@@

For gravitational waves passing through the test masses, the fractional difference in arm length $\frac{\Delta L}{L}$ is the strain $h$ (a quantity that measure the `strength' of gravitational waves) of the waves. Mathematically, the strain $h(t) = \frac{\Delta L(t)}{L}$ can also be expressed as a linear combination of its cross and plus polarizations, as follows \cite{JSTORLIGO}:
\begin{equation}
h(t) = F_\times h_\times (t) + F_+ h_+ (t)
\end{equation}
\subsubsection{LIGO: History and Prospects}

As briefly mentioned previously, LIGO (short for Laser Interferometer Gravitational-wave Observatory) is a large-scale scientific collaboration between the California Institute of Technology and the Massachusetts Institute of Technology designed to detect gravitational waves, thereby verifying Einstein's Theory of General Relativity.

In 1987, the Caltech/MIT LIGO team submitted a joint proposal to the US National Science Foundation to build two LIGO sites with the current technology; then, the initial sites would be followed by the construction of more advanced interferometers with higher probabilities of detecting gravitational waves. In 1990, the National Science Foundation approved the proposal; a year after that, Congress appropriated funding for the construction project. In 1992, NSF and Caltech signed a cooperative agreement for the management of LIGO, choosing Hanford, Washington and Livingston, Louisiana as the first sites for LIGO’s interferometers. The reason for building two sites, instead of one, is that if one interferometer detects any promising data, the other one has to as well in order for the data to be considered reliable for any detections.

In 1997, LIGO Director Barry Barish decided to split the LIGO Project into two entities:

\begin{enumerate}[nolistsep]
\item LIGO Laboratory at Caltech,
MIT, Hanford and Livingston which is responsible for LIGO operations and advanced interferometer
R\&D.
\item The LIGO Scientific Collaboration (LSC) which is responsible for coordinating LIGO's  scientific research and data analysis, and for expanding LIGO to scientists beyond Caltech and MIT.
\end{enumerate}

The two observatories were completed in the fall of 1999, with the initial interferometers up and working from 2002-2010. During this time, there were a series of six ``science runs'' called S1, S2,..., S6. LIGO first achieved its promised design sensitivity in science run S5 (2005-2007). After S5, LIGO received a series of enhancements, called ``eLIGO'', improving its performance by 20\% \cite{jstortechnical}. Then, to further upgrade LIGO, another construction project called Advanced LIGO (aLIGO) began in 2008, two years before initial LIGO was retired. Although initial LIGO and eLIGO did not successfully detect any gravitational waves, advanced LIGO did.

The construction of advanced LIGO at both sites (Hanford and Livingston) was completed in 2015. Its first run was  called O1 spanning 3 months in the fall of 2015 \cite{jstortechnical}. Very fortunately, on September 14, 2015 at 09:50:45 UTC, aLIGO detected gravitational waves from a binary merger of two black holes with masses 29 and 36 times our Sun \cite{abbott}. This event was name GW150914 (GW stands for gravitational waves and the numbers represent the date) The results were subsequently published in the \textit{Physical Review Letters} on 11 February 2016, winning LIGO's pioneers (Kip Thorne, Rainer Weiss, Barry Barish) a Nobel Prize in Physics in 2017. About three months later, on December 26, 2015 at 03:38:53 UTC, another event was detected; this time, it is from a 22-solar-mass (22 times the mass of our Sun) Binary Black Hole Coalescence \cite{GW151226}.

Following O1, O2 ran from 30 November 2016 to 25 August 2017, with Livingston exceeding O1's sensitivity by 15–25\% sensitivity \cite{collaboration2019open}, and with Hanford's sensitivity remaining around O1's. During LIGO's O2 run, another detection was confirmed as gravitational waves. On January 4, 2017 at 10:11:58 UTC, a binary black hole coalescence of 50-solar-mass was detected \cite{GW170104}. The third observing run, O3, was started on April 1, 2019 and ended on March 23, 2020 \cite{clavin_abazorius_conti_2019}.

Not only does LIGO and its detection of gravitational waves provide provide experimental confirmations of Einstein's Theory of General Relativity, it also opens entire new windows of application in other fields such as astronomy and fundamental physics. For example, gravitational waves provide a astronomers with new tools of discovering cosmological events for which traditional `electromagnetic means' cannot be used, such as black holes.    \cite{JSTORDetecting} Furthermore, gravitational waves can also contribute to pressing questions in fundamental physics such as the hierarchy problem, singularities and vacuum energy \cite{Barack_2019}.

@@

Since we are working with discrete-valued sequences, performing a DFT would give rise to what we call \textit{spectral leakages}. When the Fourier Transform is applied onto a finite, non-periodic data segment (i.e.\ Virgo/LIGO data), false frequency amplitudes might occur. The reason for this is that since the transformation is discrete in nature, sudden sharp changes at frequency bin edges when the data are not periodic can falsely create high amplitudes. Thus, to minimize this, we always apply a windowing function to data – tapering the data segment's ends, making it periodic – whenever the use of the DFT is required (note that windowing does not decrease spectral leakage, but rather re-distributes the leakage effect so that it causes the least harm). In this paper, I used a computer algorithm that implements the Discrete Fourier Transform called the Fast Fourier Transform (FFT) which is included within the Python PyCBC library \cite{numpy}.

@@

\subsubsection{Noise characteristics}
Being able to reach a strain sensitivity of $10^{-23}/\sqrt{\text{Hz}}$ \cite{sensitivity} means Virgo/LIGO's interferometers are extremely sensitive to any changes in the arm lengths – which can be caused by distorted space-time due to gravitational waves, or by vibrations that move the test masses. Thus, recorded signal $s(t)$ includes not only gravitational waves, but also unwanted vibrations from other sources as well \cite{ultimate}. Any unwanted noise, even on the smallest of scales, can still significantly interfere with data recorded by Virgo/LIGO's interferometers. Virgo/LIGO noise comes from many sources, but some examples include seismic noise, thermal noise, and shot noise.  \cite{blair_howell_ju_zhao_2012}.

Seismic noise is noise induced by the motion of the Earth's surface. It is  most prevalent in lower frequencies and can be caused by a multitude of factors, low-level earthquakes, ocean waves, and even human activities \cite{blair_howell_ju_zhao_2012}.

LIGO's mirror's thermal noise falls broadly into two categories of sources \cite{rao_libbrecht_2003}:
\begin{enumerate}[nolistsep]
\item Intrinsic thermal noise, induced by thermal forcing from internal fluctuations, causing fluctuations in mirror material properties such as length and index of refraction.
\item Extrinsic thermal noise, induced by externally imposed temperature variations that drive thermal fluctuations.One example include mirrors absorbing heat from the laser beams, thereby causing length changes by thermal expansion.
\end{enumerate}

At frequencies above 100Hz, the most prominent noise is called \textit{shot noise}. Shot noise is caused by the quantum nature of light, creating randomness in the arrival times of photons on the beam splitter.

In order to extract buried gravitational wave signals from noise, we must first understand the statistical properties of the noise. Noise is said to be stationary if its statistical properties do not significantly change over time. Another important noise characteristic is its Gaussianity. Given \textit{`Gaussian'} noise with a time-series function $n(t)$, both the real and imaginary components of its Fourier Transform $\widetilde{n}(f)$ follow the normal distribution \cite{yamamoto}; that is, $\Re[\widetilde{n}(f)]$ and  $\Im{[\widetilde{n}(f)]}$ have a probability density function (PDF) of:
\begin{equation}
    p(x) = \frac{1}{\sigma\sqrt{2\pi}}\exp{\left[-\frac{1}{2}\frac{(x-\mu)^2}{\sigma^2}\right]}
\end{equation}
where $\sigma^2$ is the sample variance and the mean $\mu = 0$. By the Central Limit Theorem in statistics, a set of random noise sample naturally tends to the Gaussian distribution if sufficiently large data samples were taken \cite{jaranowski2007gravitationalwave}. Nonetheless, real Virgo/LIGO noise can still deviate from this theoretical normality. Aside from the Gaussian distribution, noise can also follow a variety of different probability density distributions such as the Poisson distribution or the Laplace distribution. Although it is not common in real-life Virgo/LIGO operations that these noise models are present, this paper still considers the case of Laplacian noise for theoretical analysis purposes. Furthermore, there also exists abrupt, transient noise sources of, sometimes, unknown origins called glitches \cite{ultimate}. Glitches can be grouped into a variety of classifications according Gravity Spy's \textit{`Field Guide'} \cite{gravityspy}. In this paper, we only primarily discuss five types of glitches: \textit{Extremely Loud}, \textit{Scattered Light}, \textit{Power Line}, \textit{Whistle}, and \textit{Blip}. \textit{Extremely Loud} is a broad classification for major detector disturbances such as photodiode saturation, causing loud and high energy transient noise bursts. \textit{Scattered Light} glitches are caused by malfunctions in the optical mechanisms of Virgo/LIGO; they look like upward humps and are particularly harmful in searching for binary inspirals \cite{gravityspy}. The third classification of glitches is called \textit{Power Line}; these glitches are caused by electrical equipments such as air compressors that use alternating current (AC). The fourth classification is called \textit{Whistle} glitches; they are usually reverse V-shaped and are caused by radio signals \cite{gravityspy}. Finally, \textit{Blip} glitches are very short duration ($\approx 40$ms) noise bursts at frequencies between 30 and 500 Hz. As of currently, there are no known causes and thus \textit{Blip} glitches are the most damaging to the detection of gravitational waves \cite{gravityspy}. It is worth pointing out that over short time intervals, though, Virgo/LIGO noise can still be reasonably approximated to be Gaussian and stationary \cite{collaboration2019open}.

\begin{widetext}
\begin{figure}[t]
\includegraphics[width = .9\textwidth]{gaussian -21 template 1.png}
\caption{The resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Gaussian noise scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 20.0$}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 gaussian zoomed.png}
\caption{The resulting $\chi^2_r$ statistic for perfectly Gaussian noise scaled down by a factor of $10^{-21}$. A strong dip toward unity is clearly seen at the expected time.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 gaussian.png}
\caption{The resulting $\chi^2_r$ statistic for perfectly Gaussian noise scaled down by a factor of $10^{-21}$, zoomed in to just the two-second interval between the time of expected signal.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{laplacian b=1 template 1.png}
\caption{The resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Laplacian noise with $b=1$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 12.6$.}

@@

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=1 zoomed.png}
\caption{The resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=1$ scaled down by a factor of $10^{-21}$. A strong dip toward unity is seen with climbing values just around it.}
\centering
\end{figure} 
\end{widetext}

@@

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=1.png}
\caption{The resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=1$ scaled down by a factor of $10^{-21}$. The graph is zoomed in to show just the two-second time interval around the time of expected signal. A strong dip toward unity is seen.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{laplacian b=2 template 1.png}
\caption{The resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Laplacian noise with $b=2$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 35.0$s with SNR $\rho = 6.1$.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=2 zoomed.png}
\caption{The resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=2$ scaled down by a factor of $10^{-21}$. A small dip toward unity is seen.}
\centering
\end{figure} 
\end{widetext}
\subsubsection{Matched filtering}
The primary goal of Virgo/LIGO data analysis is to determine if there is gravitational wave signal buried in the noise. There are many ways that this can be accomplished; however, the most efficient and effective method is called \textit{matched filtering}, especially in cases where noise is both Gaussian and stationary \cite{helstrom_1975}. Matched filtering allows the recorded data stream to be searched for the presence of a known gravitational wave template with varying arrival time and phase offset \cite{surf}. Matched filtering requires three components: the recorded data stream $s(t)$, the \textit{power spectral density (PSD)} function $S_n(t)$ of the noise, and a known waveform template $h(t)$ of the gravitational wave. (Note that the PSD is essentially a measure of how power is distributed at different frequencies of the signal; detailed mathematical descriptions can be found in \cite{psd}.) Given these quantities, the complex-valued output of matched filtering at time $t=t_0$ can be computed as follows \cite{findchirp}:
\begin{equation}
    z(t_0) = 4\int_0^\infty\frac{\widetilde{s}(f)\widetilde{h}^*(f)}{S_n(f)}e^{2\pi\iu ft_0}df
\end{equation}
where $\widetilde{h}^*(f)$ denotes the complex conjugate of the Fourier Transform of the template strain $h(t)$. Notice that the existence of $S_n(f)$ in the denominator essentially down-weighs the frequencies where the noise is loud and does the opposite where the noise is weak \cite{ultimate}. Given the complex matched filter output $z(t)$, we can compute a test statistic value $\rho$ called the \textit{signal-to-noise ratio} as follows \cite{findchirp}: 
\begin{equation}
   \rho (t) = \frac{|z(t)|}{\sigma}
\end{equation}
where $\sigma$ is a normalization factor is computed from \cite{findchirp}
\begin{equation}
   \sigma^2 =  4\int_0^\infty\frac{|\widetilde{h}(f)|^2}{S_n(f)}df
\end{equation}

@@

In the presence of non-Gaussian and/or non-stationary noise and/or glitches, matched filtering does not perform as well as expected, since the filter itself is derived from Gaussian and stationary noise. A detailed mathematical derivation of the matched filter from a stationary, Gaussian noise model can be found in \textit{Brown (2007)} \cite{brown2007searching}.

There are several ways of incorporating matched filtering into a computer program that analyzes LIGO's signal. In this paper, I only make use of the matched filtering algorithm implemented in the open-source Python software package PyCBC \cite{pycbc}. This algorithm is based on the FINDCHIRP algorithm presented above and in \cite{findchirp}.
\begin{widetext}
\begin{figure}[t]
\includegraphics[width = .9\textwidth]{chi2 laplace b=2.png}
\caption{The resulting $\chi^2_r$ statistic for perfectly Laplacian noise with $b=2$ scaled down by a factor of $10^{-21}$. The graph is zoomed in to show just the two-second time interval around the time of expected signal. A relatively weak dip toward unity is seen.}
\centering
\end{figure}

\begin{figure}[t]
\includegraphics[width = .9\textwidth]{laplacian b=3 template 1.png}
\caption{The resulting cropped SNR $\rho(t)$ after matched filtering was applied to recover the template signal from perfectly Laplacian noise with $b=3$, scaled down by a factor of $10^{-21}$. A signal peak was detected at time $t = 21.7$s with SNR $\rho = 4.6$.}
\centering
\end{figure}
\end{widetext}
